 

### 2021-01-27 vertical-thermal-review.Rmd
review of "The Biophysical Environment"

The biophysical environment, here defined to include the physical structure of the forest canopy (from understory to the top of the canopy) and associated microclimates, is in large part determined by the structure of the forest itself. 
We supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S. National Ecological Observatory Network [NEON; Appendix `r appendix_NEON_methods`; Schimel et al. 2007].

Canopy foliage, which varies across forest types and seasonality, strongly shapes understory light, wind, etc.
Forest canopies have a buffering effect on multiple aspects of the understory climate (Fig. `r fig_NEON_vertical`).

- (*@bonan_ecological_2016 reviews this and points to appropriate references*)

*(paragraph on leaf area distribution, including figure ([GitHub issue #15](https://github.com/EcoClimLab/vertical-thermal-review/issues/15))*
The vertical structure and corresponding microenvironments of forest canopies vary seasonally (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019) and following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season, but excluding periods when no leaves are present in deciduous forests.

Light conditions, specifically changes in photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [@koike_leaf_2001]. Generally, upper canopies in dense forests are exposed to greater PAR values and limit light penetration to canopy interiors, lower canopy layers, and the understory. Shaded regions experience lower values of PAR with different light quality [in ranges of blue, orange and red]. The light gradient is more pronounced under broadleaf forests than needleleaf forests [@lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b]
(Fig. `r fig_NEON_vertical`).

@@

**Wind speeds are also higher at the top of the canopy (Fig. `r fig_NEON_vertical`).**
*(fill in some specifics from the NEON analysis)*
Consistent with the results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019]

Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a]. Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011]. In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017]. Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy? [ref]

**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019], and recent evidence shows a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a] and in the northwestern United States [@davis_microclimatic_2019]. Intact tropical forests have also been show to exhibit lower peak daytime $T_{air}$ than neighboring tree crop agricultural plantations (S. R. Hardwick et al., The relationship between leaf area index and microclimate in tropical forest and oil palm plantation: Forest disturbance drives changes in microclimate. Agric. Meteorol. 201, 187–195 (2015)). Even though canopy cover buffers maximum air temperatures in the understory, inevitable increases in understory temperatures continue to be observed as a result of global temperature increases. However, the buffering by canopy cover smooths this warming in the understory, thereby exhibiting an understory temperature lag [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020;@rollinsonClimateSensitivityUnderstory]

@@

Minimum $T_{air}$ is also buffered by forest canopies under some conditions.
Higher (Elsa: or lower?) minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].(Martijn: should we refer to mechanisms at this point? (e.g., greater radiative heat loss in exposed areas resulting in lower Tmin))
However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].

@@

**Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**
- DETAILS - REFS
- @mcgregor_tree_nodate

@@

**Finally, carbon dioxide (CO~2~) concentrations tend to be higher in the understory.** 
	Higher in the understory, particularly at dusk [@koike_leaf_2001]. Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999]. Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [because of lower photosynthetic rate in shade leaves?] [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]

Given that differences are small during the day when photosynthesis is occuring, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentrations are unlikely to have much of an effect on the energy balance and metabolism of leaves across the forest vertical gradient.

If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO~2~ emitted at the forest floor (i.e. from soils and decaying organic matter and leaf litter) could increase the vertical CO~2~ gradient in the future, but this would still only be a short-term effect in the early morning, as the CO~2~ would quickly diffuse through the canopy. Furthermore, increases in mortality of big trees (e.g. @senfCanopyMortalityHas2018a) would increase canopy roughness and prevent build-up of CO~2~ concentration gradients.
(Elsa: see these papers on the long-term Harvard Forest soil warming experiments. CO2 emissions increased, then decreased, then increased again. It's thought that the changes in soil CO2 emissions over time were linked to turnover in microbial community composition.
- DeAngelis, K.M., Pold, G., Topçuoğlu, B.D., van Diepen, L.T., Varney, R.M., Blanchard, J.L., Melillo, J. and Frey, S.D., 2015. Long-term forest soil warming alters microbial communities in temperate forest soils. Frontiers in microbiology, 6, p.104.
- Melillo, J.M., Steudler, P.A., Aber, J.D., Newkirk, K., Lux, H., Bowles, F.P., Catricala, C., Magill, A., Ahrens, T. and Morrisseau, S., 2002. Soil warming and carbon-cycle feedbacks to the climate system. Science, 298(5601), pp.2173-2176.
- Melillo, J.M., Frey, S.D., DeAngelis, K.M., Werner, W.J., Bernard, M.J., Bowles, F.P., Pold, G., Knorr, M.A. and Grandy, A.S., 2017. Long-term pattern and magnitude of soil carbon feedback to the climate system in a warming world. Science, 358(6359), pp.101-105.) 
- Nottingham, A.T., Meir, P., Velasquez, E. and Turner, B.L., 2020. Soil carbon loss by experimental warming in a tropical forest. Nature, 584(7820), pp.234-237.

**The strength of this buffering varies across forests, influenced by both forest characteristics (Fig. `r fig_NEON_differences`) and the biophysical environment.**
First, buffering increases with canopy cover and water availability.
* greater cover --> lower max T and VPD, higher minT  [@davis_microclimatic_2019]
* greater cover -->  lower max T and light availability [@zellweger_seasonal_2019]
* [@thom_effects_2020; Tymen et al. 2017 ; Jucker et al. 2018 ; Hardwick et al. 2015] 
The omega factor (McNaughton and Jarvis, 1983; Jarvis & McNaughton 1986) describes the degree of aerodynamic coupling between the vegetation and the atmospheric boundary layer; the lower the value of omega, the greater the coupling.
"I wonder whether it would be useful to introduce such 'formal' terms used to describe what is essentially the same as buffering effect of canopies.
I'd have to read up some more to see if there are papers showing what sort of canopy characteristics influence omega.

@@

![**Figure `r fig_NEON_differences`. Role of canopy permeability in shaping temperature buffering, from NEON data.** Current placeholder figure is Nidhi's first rough analysis. See [issue 31](https://github.com/EcoClimLab/vertical-thermal-review/issues/31). ](NEON_height_profiles/figures/differences_across_forests_temp.png)

>>>>

**The biophysical environment across the vertical gradient from the forest floor to the the top of the canopy is in large part determined by the structure of the forest itself.** 
Here, we define the biophysical environment to include the physical structure of the vegetation and associated physical conditions.
In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S. National Ecological Observatory Network [NEON; Appendix `r appendix_NEON_methods`; Schimel et al. 2007].

### Foliage density
**Canopy foliage, which is vertically structured in correspondence with the size structure of trees in the forest, acts as the primary physical barrier between the atmosphere and the forest floor, producing a buffering effect on multiple aspects of the understory conditions (Fig. `r fig_NEON_vertical`).**
Leaf area density.... *description of most common vertical pattern (@bonan_ecological_2016 reviews this and points to appropriate references)*
Vertical profiles in leaf area density vary across forests, being... *summarize some key differences across forest types* (REFS), and are altered (*HOW?*) following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
They also vary seasonally, ... (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
*(can we say that there is more loss of leaves higher in the profile during cold/ drought dormant seasons? )*
In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season, but excluding periods when no leaves are present in deciduous forests.

### Light environments
**Light conditions, specifically changes in proportion incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, upper canopies in dense forests are exposed to greater PAR values and limit light penetration to canopy interiors, lower canopy layers, and the understory. Shaded regions experience lower values of PAR with different light quality [in ranges of blue, orange and red]. The light gradient is more pronounced under broadleaf forests than needleleaf forests [@lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b]
(Fig. `r fig_NEON_vertical`).

@@

### Wind speeds 
**Wind speeds are also higher at the top of the canopy (Fig. `r fig_NEON_vertical`).**
*(fill in some specifics from the NEON analysis)*
Consistent with the results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019]

Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a]. *(mention boundary layer conductance here, but move the discussion of leaf temperature to the T_leaf section )* ~~Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011]. In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017]. Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy? [ref] ~~

### Air temperature 
**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019], and recent evidence shows a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a] and in the northwestern United States [@davis_microclimatic_2019]. Intact tropical forests have also been show to exhibit lower peak daytime $T_{air}$ than neighboring tree crop agricultural plantations (S. R. Hardwick et al., The relationship between leaf area index and microclimate in tropical forest and oil palm plantation: Forest disturbance drives changes in microclimate. Agric. Meteorol. 201, 187–195 (2015)). Even though canopy cover buffers maximum air temperatures in the understory, inevitable increases in understory temperatures continue to be observed as a result of global temperature increases. However, the buffering by canopy cover smooths this warming in the understory, thereby exhibiting an understory temperature lag [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020;@rollinsonClimateSensitivityUnderstory]

@@

Minimum $T_{air}$ is also buffered by forest canopies under some conditions.
Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].(Martijn: should we refer to mechanisms at this point? (e.g., greater radiative heat loss in exposed areas resulting in lower Tmin))
However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].

@@

![**Figure `r fig_NEON_differences`. Relationship of PAR differences between canopy and understory to air temperature buffering (a) across sites and (b) through time within a single site.** From NEON data. Current placeholder figure is Nidhi's first rough analysis. See [issue 31](https://github.com/EcoClimLab/vertical-thermal-review/issues/31). ](NEON_height_profiles/figures/differences_across_forests_temp.png)

### Humidity and evaporative demand
**Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**
- DETAILS - REFS
- @mcgregor_tree_nodate

@@

### Carbon dioxide
**Carbon dioxide (CO~2~) concentrations tend to be higher in the understory, primarily at night.** 
	Higher in the understory, particularly at dusk [@koike_leaf_2001]. Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999]. Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [because of lower photosynthetic rate in shade leaves?] [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]
Given that differences are small during the day when photosynthesis is occurring, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentrations are unlikely to have much of an effect on the energy balance and metabolism of leaves across the forest vertical gradient.
If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO~2~ emitted at the forest floor (i.e. from soils and decaying organic matter and leaf litter) could increase the vertical CO~2~ gradient in the future, but this would still only be a short-term effect in the early morning, as the CO~2~ would quickly diffuse through the canopy. 
~~Furthermore, increases in mortality of big trees (e.g. @senfCanopyMortalityHas2018a) would increase canopy roughness and prevent build-up of CO~2~ concentration gradients.~~ (*but make the point about big tree mort elsewhere*)

(*keep discussion of CO2 to what's in the paragraph above. Definitely don't want to go beyond a brief mention of warming effects on soil respiration*): (Elsa: see these papers on the long-term Harvard Forest soil warming experiments. CO2 emissions increased, then decreased, then increased again. It's thought that the changes in soil CO2 emissions over time were linked to turnover in microbial community composition.
- DeAngelis, K.M., Pold, G., Topçuoğlu, B.D., van Diepen, L.T., Varney, R.M., Blanchard, J.L., Melillo, J. and Frey, S.D., 2015. Long-term forest soil warming alters microbial communities in temperate forest soils. Frontiers in microbiology, 6, p.104.
- Melillo, J.M., Steudler, P.A., Aber, J.D., Newkirk, K., Lux, H., Bowles, F.P., Catricala, C., Magill, A., Ahrens, T. and Morrisseau, S., 2002. Soil warming and carbon-cycle feedbacks to the climate system. Science, 298(5601), pp.2173-2176.
- Melillo, J.M., Frey, S.D., DeAngelis, K.M., Werner, W.J., Bernard, M.J., Bowles, F.P., Pold, G., Knorr, M.A. and Grandy, A.S., 2017. Long-term pattern and magnitude of soil carbon feedback to the climate system in a warming world. Science, 358(6359), pp.101-105.) 
- Nottingham, A.T., Meir, P., Velasquez, E. and Turner, B.L., 2020. Soil carbon loss by experimental warming in a tropical forest. Nature, 584(7820), pp.234-237.

*ALL OF THE FOLLOWING CONTENT SHOULD BE WORKED INTO RELEVANT PARAGRAPHS EARLIER IN THIS SECTION:*

The strength of this buffering varies across forests, influenced by both forest characteristics (Fig. `r fig_NEON_differences`) and the biophysical environment.
First, buffering increases with canopy cover and water availability.
* greater cover --> lower max T and VPD, higher minT  [@davis_microclimatic_2019]
* greater cover -->  lower max T and light availability [@zellweger_seasonal_2019]
* [@thom_effects_2020; Tymen et al. 2017 ; Jucker et al. 2018 ; Hardwick et al. 2015] 
The omega factor (McNaughton and Jarvis, 1983; Jarvis & McNaughton 1986) describes the degree of aerodynamic coupling between the vegetation and the atmospheric boundary layer; the lower the value of omega, the greater the coupling.
"I wonder whether it would be useful to introduce such 'formal' terms used to describe what is essentially the same as buffering effect of canopies.
I'd have to read up some more to see if there are papers showing what sort of canopy characteristics influence omega.

@@
